Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-20 paper/paper.tex
First draft, new intro, massaged related work.

This paper presents the first Coq formalisation of \emph{hylomorphisms} allowing for the
mechanisation of all recognised recursive algorithms. The key
contribution of this paper is that this formalisation is fully \emph{axiom-free}
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted OCaml code for the programs formalised in our
framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
the machine-checked proofs of the laws of hylomorphisms to do program
optimisations.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
reasoning about program equivalences, transformations, and
optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

There is not much tool support to use program calculation techniques. In fact,
uses of program calculation are done by  pen-and-paper proofs, and then
translating the result to specific instances of recursion schemes, implemented
in a programming language (generally Haskell).  There are some examples of
implementations of program calculation techniques, but these implementations are
scarce, not up to date, and not verified in a proof assistant. For example,
Cunha et al.~\cite{DrHylo} automated program calculation techniques by a custom
Haskell implementation.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the
existing work only mechanises a narrow subset of the know recursion schemes, or
do not prove their algebraic
laws~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
most of the mechanisations of generic recursion schemes cover \emph{maps} and
\emph{folds}, and few authors focus implement \emph{unfolds}. Omitting
\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
and prevent them from being able to mechanise the most general recursion scheme:
a divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
schemes are instances of hylomorphisms}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant, together with their algebraic laws.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code, and it is not well-suited for
program calculation. This is unfortunate since code extration is what allows the
execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al's approach, extraction (1) does not produce code with the same recursive
structure  as existing manual implementations; and (2) leads to unsafe casts
like \ocaml{Obj.magic} in the generated code. This latter is also problematic in
that, for higher-order programs, simple interoperations can lead to incorrect
behaviour or even segfaults~\cite{forster:hal-04329663} and, moreover, it
invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.

@@

\section{Related Work}
\label{sec:related-work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques and recursion schemes in Coq, and
discuss other approaches to fusion and to termination of nonstructural
recursion.
% and recursion schemes in Coq
% (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
% encoding of recursion schemes to other forms of termination checking for
% nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
% approaches to fusion in functional programs (Section~\ref{sec:fusion}).

\subsection{Program Calculation}\label{sec:calccoq}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

@@

\subsection{Fusion Optimisation}\label{sec:fusion}

Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

@@

%\mvol{Need more general citations for program calculation.}
\subsection{Divide and Conquer Recursion}\label{sec:reccoq}

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing. This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}. The
authors justify well the use of impredicative \coq{Set} and its compatibility
with the functional extensionality axiom. In contrast, our development remains
entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
OCaml code.
% Another key difference with our
% approach is that they do not discuss what the resulting extracted code
% \emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that code extracted from their formalisation makes
heavy use of \ocaml{Obj.magic} and has a generally complex structure that may be
hard to understand and integrate with other OCaml code. Due to the great benefit
of entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

\subsection{Termination Checking of Nonstructural Recursion}\label{sec:termcheck}

% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.

%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

>>>>

This paper presents the first Coq formalisation of \emph{hylomorphisms} allowing for the
mechanisation of all recognised recursive algorithms. The key
contribution of this paper is that this formalisation is fully \emph{axiom-free}
allowing for the extraction of safe, idiomatic functional code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted functional code for the programs formalised in our
framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
the machine-checked proofs of the laws of hylomorphisms to do program
optimisations.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Structured recursion schemes are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
reasoning about program equivalences, transformations, and
optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. A lot of program calculations have this pattern:
start from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version. We will revisit a similar example
in Section~\ref{sec:sorting}.

This program calculation stems from the more general theory of hylomorphisms.  A
hylomorphism is a structured recursion scheme representing the idea of a
divide-and-conquer algorithm where, first, the problem is split into smaller
subproblems,  (divide), then subsolutions are computed recursively, and,
finally, these subsolutions are put together to compute the original solution
(conquer). This general pattern of recursion induces a reasoning principle which
is called fusion law which has been exemplified in the previous paragraph.

Hylomorphism are very powerful non-structural recursive algorithms that can be
used to implement structural recursion, as well as many other known recursion
schemes such as mutual recursion, accumulators, primitive recursion and
course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
comes at a price. Since hylomorphisms are not necessarily structural they need
extra-care is needed when encoding these in a type theory. In particular, the
divide part of the algorithm must be \emph{recursive} that is the input needs to
be split into smaller parts (not necessarily following the structure of the
input data).

Implementations of recursion schemes currently focus on non-terminating
languages (e.g. Haskell) and they would benefit from a type-theoretical
formalisation which leverages the underlying prover's logic to ensure
correctness of definitions.  While existing mechanisations do exist they either
do not cover a large subset of the recursion schemes or are not suitable for
program calculation or code extraction (see Related
Work~\ref{sec:related-work}).

In this work we mechanise hylomorphisms thus reaping the benefits of their
generality and their algebraic properties. We do this in a way which allows the
user to write specifications of their code, reason about program optimisations
and eventually be able to extract humanly readable functional code which can be
run which means it cannot contain unsafe casts. These are, in particular,
problematic  for higher-order programs as simple interoperations with them can
lead to incorrect behaviour or even segmentation
faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

To retain the hylomorphism's generality we encode data structures using
polynomial functors and to ensure our extracted code is runnable we make a point
to remain axiom-free and avoid indexed types in the definition of our recursion
schemes.

@@

\section{Related Work}
\label{sec:related-work}

%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
We compare our work with other
encodings of program calculation techniques and recursion schemes in Coq, and
discuss other approaches to fusion and to termination of nonstructural
recursion.
% and recursion schemes in Coq
% (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
% encoding of recursion schemes to other forms of termination checking for
% nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
% approaches to fusion in functional programs (Section~\ref{sec:fusion}).

% \textbf{Program Calculation}.
% \label{sec:calccoq}
Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

@@

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

@@

%\mvol{Need more general citations for program calculation.}
% \textbf{Divide and Conquer Recursion}.%
% \label{sec:reccoq}
Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing. This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}. The
authors justify well the use of impredicative \coq{Set} and its compatibility
with the functional extensionality axiom. In contrast, our development remains
entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
OCaml code.
% Another key difference with our
% approach is that they do not discuss what the resulting extracted code
% \emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that code extracted from their formalisation makes
heavy use of unsafe casts and has a generally complex structure that may be
hard to understand and integrate with other OCaml code. Due to the great benefit
of entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.
%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.
